Greenhouse Locoweed VOC Collection Report

### 2024-05-13 qmd/projects/msu/locoweed/gh-locoweed-report.qmd
commit 5/13 2024

</body>
</html>

*Note: This project is still a work in progress.*

# Introduction
Astragalus and Oxytropix species, part of the Legume Family, are found in terrestrial ecoysstems globally, including the western regions of North America. Many of these species are both nontoxic and important forages. However, several species are known to be toxic to both livestock and wildlife. *Oxytropis sericea* Nutt can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is associated with locoweed plants, specifically those of the genus *Oxytropis*. Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities [@locoweed]. In this study, we investigated the potential differences in volatile organic compound (VOC) emissions between specimens that contained the commensal endophyte *Alternaria oxytropis*, and those that did not.

@@

# Methodology
## Data Collection

### Greenhouse Volatile Collection
Samples of each plant were first sent for analysis to mystery lab to see if endophyte was present or not. 
Volatile compounds were collected from healthy *(size/stage)* locoweed plants. Plants were paired based on size similarity, with one being E+ and the other E-. Three sets of pairs (twelve plants) were used for each volatile collection event. Collection was undergone for a period of six hours from 09:00 to 15:00.

<div class="box">
<!-- plant pictures -->
```{r plant pictures, echo = F, eval = F}

@@

```
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Image Gallery</title>
  <style>
    .gallery-container {
      display: flex;
      justify-content: center;
      align-items: center;
      height: 100%;
    }
    .gallery {
      display: grid;
      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
      grid-gap: 10px;
    }
    .gallery img {
      width: 100%;
      height: 100%;
      object-fit: cover; /* Ensure images cover the entire space */
      cursor: pointer;
      transition: transform 0.3s ease;
    }
    .gallery img:hover {
      transform: scale(1.05); /* Slightly increase size on hover */
    }
    .expanded-container {
      position: fixed;
      top: 0;
      left: 0;
      width: 100%;
      height: 100%;
      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
      display: none; /* Initially hidden */
      justify-content: center;
      align-items: center;
      z-index: 9999;
    }
    .expanded {
      max-width: 80%;
      max-height: 80%;
      cursor: pointer;
    }
    p.gallery-caption {
      padding-top: 5px;
      text-align: center;
      font-style: italic;
      color: #8B8989;
    }
  </style>
</head>
<body>

@@

</body>
</html>
```
</div>

@@

## Data Analysis
The data analysis procedure for each collection type was the same. Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling. The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide]. VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan]. This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017]. Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr]. I selected a subset of compounds to ues for each analysis. These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.

# RESULTS
## Visual Comparisons
<!-- LOAD DATA -->
```{r gh RAW 22, echo = F}
raw.gh.22 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2022\ Spring\ Greenhouse/Spring\ 2022\ Locoweed/Spring\ 2022\ GH\ Locoweed\ RESULTS.xlsx", sheet = "RAW2") %>%
  filter(remove == "no")
```
```{r gh RAW 23, echo = F}
raw.gh.23 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2023\ Spring\ Greenhouse/Spring\ GH\ 23\ Locoweed\ Volatiles.xlsx", sheet = "RAW_PROC")

```
<!-- CLEAN DATA -->
```{r gh CLEAN 22, echo = F}
clean.gh.22 <-
raw.gh.22 %>%
  # filter(compound %in% c("heptanal", "sixmethyl", "octanal",
  #                        "z3hex", "dlimonene", "nonanal", 
  #                        "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
  #             values_from = "ngghr") %>%
  left_join(raw.gh.22 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2022")
```
```{r gh CLEAN 23, echo = F}
clean.gh.23 <-
raw.gh.23 %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
              # values_from = "ngghr") %>%
  left_join(raw.gh.23 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2023") 
```
<!-- COMOPUNDS -->
```{r gh compounds, echo = F, results = F}
compounds22 <-
raw.gh.22 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds23 <-
raw.gh.23 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds22
compounds23 %>%
  print(n = 100)
gh.compounds <- as.list(intersect(compounds22, compounds23))$compound
```
<!-- CLR -->
```{r gh CLR, echo = F}
gh.joined.data <-
clean.gh.22 %>%
  dplyr::select(plant.id, compound, ngghr, swa, year, pair) %>%
  rbind(clean.gh.23 %>%
          dplyr::select(plant.id, compound, ngghr, swa, year, pair)) %>%
  distinct() %>%
  filter(compound %in% gh.compounds) %>%
  na.omit()

@@

### Box Plot
```{r gh boxplot comparison, echo = F, fig.align = 'center', fig.width = 10}
gh.joined.data.clr %>%
  # pivot to make all compounds in one column
  pivot_longer(-c(plant.id, year, swa, pair),
               names_to = "compound",
               values_to = "ngghr") %>%
  # ggplot code
  ggplot(aes(compound, ngghr, fill = swa)) +
  geom_boxplot(position = position_dodge(width = 0.9),
                 outlier.color = NA,
                 alpha = 0.5) +
  geom_jitter(aes(compound, ngghr),
              alpha = 0.3,
              position = position_jitterdodge(jitter.width = 0.2,
                                              dodge.width = .8),
              show.legend = FALSE,
              # pch = 21, 
              stroke = 0, 
              size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.background = element_rect(color = "black"),
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        panel.grid.major.x = element_line(color = "#F2F2F2")) +
  labs(x = element_blank(),
       y = "VOC quantity (CLR(ng/g/hr))",
       title = "2022/23 Greenhouse Locoweed Volatile Profile Comparison") +
  scale_fill_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
```
:::{.callout-note}
I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed. As we can see, there isn't a tremendous amount of obvious variance between E+ and E- samples. There is, perhaps, more of a difference at first glance than what we saw in our [field volatile analysis](field-locoweed-report.qmd#visual-comparisons) bar plots.  
:::

### PCA
Principal Component Analysis (PCA) is a dimensionality reduction technique used to transform high-dimensional data into a lower-dimensional representation while preserving most of the original variability. It identifies the directions (principal components) that capture the maximum variance in the data and projects the data onto these components.

@@

## Check Assumptions  
The statistical analyses we can perform are greatly impacted by our data's ability to meet statistical assumptions. One of the major assumptions here is normality. We are going to test this assumption using the `MVN` package.

@@

As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation. 
</div>

:::{.callout-note}
**A quick note here.**  
Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance). Despite the similarity of the data to ours in many instances, MANOVA is used even though the data being analyzed violates the normality assumption. Since our data clearly violates this assumption, we have decided against using this method.
:::

## Kruskal-Wallis Test
The [Kruskal-Wallis](https://en.wikipedia.org/wiki/Kruskal–Wallis_test) test is a non-parametric statistical test used to determine if there are differences between two or more independent groups. It assesses whether the distributions of scores across groups are the same or different, without assuming the data follow a normal distribution, making it suitable for ordinal or non-normally distributed data.

@@

### PERMANOVA  
As described above, a [PERMANOVA](https://en.wikipedia.org/wiki/Permutational_analysis_of_variance) (Permutational Multivariate Analysis of Variance) uses permutation testing to generate many different rearrangements of the data and assess if the observed differences are statistically significant [@anderson2017]. This method is important as it considers all of the volatile profiles overall when generating a p-value.

@@

```{r PERMANOVA}
set.seed(1)
gh.permanova.output <-
adonis2(gh.permanova.data ~ gh.data.per.meta$swa, 
        strata = interaction(gh.data.per.meta$pair,
                             gh.data.per.meta$year),
        # method = "canberra",
        permutations = 9999,
        by = "terms")
```
```{r PERMANOVA table, echo = F}
kable(as.data.frame(gh.permanova.output),
      format = "html",
      caption = "Table 1: Greenhouse Locoweed VOC collection PERMANOVA results.   gh.data.per.meta$swa is our E+/E- variable.",
      align = "lccccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f9") 
```

@@

```{r random forest}
ranger_model <- ranger(swa ~ ., 
                       data = gh.joined.data.clr %>% 
                         dplyr::select(-plant.id),
                       importance = "impurity")
```
```{r random forest formatting and plot, echo = F}
var_importance <- importance(ranger_model)

>>>>

</body>
</html>
```
*Note: This project is still a work in progress.*

# Introduction

Astragalus and Oxytropix species, part of the Legume Family, are found in terrestrial ecoysstems globally, including the western regions of North America. Many of these species are both nontoxic and important forages. However, several species are known to be toxic to both livestock and wildlife. *Oxytropis sericea* Nutt can harbor a commensal endophyte, *Alternaira oxytropos*, a fungus that is associated with locoweed plants, specifically those of the genus *Oxytropis*. Plants containing the endophyte have been found to produce swansonine, a toxic alkaloid that can be harmful to livestock when ingested in large quantities [@locoweed]. In this study, we investigated the potential differences in volatile organic compound (VOC) emissions between specimens that contained the commensal endophyte *Alternaria oxytropis*, and those that did not.

@@

# Methodology

## Data Collection

### Greenhouse Volatile Collection

Samples of each plant were first sent for analysis to mystery lab to see if endophyte was present or not. Volatile compounds were collected from healthy *(size/stage)* locoweed plants. Plants were paired based on size similarity, with one being E+ and the other E-. Three sets of pairs (twelve plants) were used for each volatile collection event. Collection was undergone for a period of six hours from 09:00 to 15:00.

::: box
<!-- plant pictures -->

```{r plant pictures, echo = F, eval = F}

@@

```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Image Gallery</title>
  <style>
    .gallery-container {
      display: flex;
      justify-content: center;
      align-items: center;
      height: 100%;
    }
    .gallery {
      display: grid;
      grid-template-columns: repeat(auto-fill, minmax(150px, 1fr));
      grid-gap: 10px;
    }
    .gallery img {
      width: 100%;
      height: 100%;
      object-fit: cover; /* Ensure images cover the entire space */
      cursor: pointer;
      transition: transform 0.3s ease;
    }
    .gallery img:hover {
      transform: scale(1.05); /* Slightly increase size on hover */
    }
    .expanded-container {
      position: fixed;
      top: 0;
      left: 0;
      width: 100%;
      height: 100%;
      background-color: rgba(0, 0, 0, 0.8); /* Semi-transparent background */
      display: none; /* Initially hidden */
      justify-content: center;
      align-items: center;
      z-index: 9999;
    }
    .expanded {
      max-width: 80%;
      max-height: 80%;
      cursor: pointer;
    }
    p.gallery-caption {
      padding-top: 5px;
      text-align: center;
      font-style: italic;
      color: #8B8989;
    }
  </style>
</head>
<body>

@@

</body>
</html>
```
:::

@@

## Data Analysis

The data analysis procedure for each collection type was the same. Statistical analyses were conducted in R (R Core Team 2024) using RStudio (version 4.2.1) using PERMANOVA, PERMDISP, and Random Forest modeling. The data was first transformed using a center log-ratio (CLR) transformation, a technique commonly used when analyzing multivariate compositional data [@chemoguide]. VOC compositions were compared between samples and treatments using a permutational multivariate analysis of variance (PERMANOVA, Canberra distance, 999 iterations) tests implemented in the vegan package (v 2.6-4) [@vegan]. This technique uses permutation testing to generate many different rearrangements of the data to assess if the observed differences are statistically significant [@anderson2017]. Data were visualized using a combination of ggplot2 and knitr [@ggplot; @knitr]. I selected a subset of compounds to ues for each analysis. These compounds were selected based on consistency of presence across samples and years, as well as overall quality of mass spectroscopy output.

# RESULTS

## Visual Comparisons

<!-- LOAD DATA -->

```{r gh RAW 22, echo = F}
raw.gh.22 <-
read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2022\ Spring\ Greenhouse/Spring\ 2022\ Locoweed/Spring\ 2022\ GH\ Locoweed\ RESULTS.xlsx", sheet = "RAW2") %>%
  filter(remove == "no")
```

```{r gh RAW 23, echo = F}
raw.gh.23 <-
  read_excel("/Users/jacksonstrand/Library/CloudStorage/OneDrive-MontanaStateUniversity/Data/Locoweed/Greenhouse/2023\ Spring\ Greenhouse/Spring\ GH\ 23\ Locoweed\ Volatiles.xlsx", sheet = "RAW_PROC")

<!-- CLEAN DATA -->

```{r gh CLEAN 22, echo = F}
clean.gh.22 <-
raw.gh.22 %>%
  # filter(compound %in% c("heptanal", "sixmethyl", "octanal",
  #                        "z3hex", "dlimonene", "nonanal", 
  #                        "decanal", "methylsalicylate", "cisthreehexiso")) %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
  #             values_from = "ngghr") %>%
  left_join(raw.gh.22 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2022")
```

```{r gh CLEAN 23, echo = F}
clean.gh.23 <-
raw.gh.23 %>%
  dplyr::select(plant.id, compound, ngghr) %>%
  group_by(plant.id, compound) %>%
  na.omit() %>%
  filter(!compound == "standard") %>%
  dplyr::reframe(ngghr = sum(ngghr)) %>%
  distinct() %>%
  pivot_wider(names_from = "compound",
              values_from = "ngghr") %>%
  pivot_longer(-plant.id,
               names_to = "compound",
               values_to = "ngghr") %>%
  replace(is.na(.), 0) %>% filter(compound != "NA") %>%
  mutate(ngghr = log(ngghr + 1.01, 2)) %>%
  # pivot_wider(names_from = "compound",
              # values_from = "ngghr") %>%
  left_join(raw.gh.23 %>% 
              dplyr::select(plant.id, swa, pair), # go back and add pairs later
            by = "plant.id",
            relationship = "many-to-many") %>%
  # replace(is.na(.), 0) %>%
  distinct() %>%
  mutate_at(vars(swa), list(factor)) %>%
  ungroup() %>%
  mutate(year = "2023") 
```

<!-- COMOPUNDS -->

```{r gh compounds, echo = F, results = F}
compounds22 <-
raw.gh.22 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds23 <-
raw.gh.23 %>%
  dplyr::select(compound) %>%
  na.omit() %>% distinct() 
compounds22
compounds23 %>%
  print(n = 100)
gh.compounds <- as.list(intersect(compounds22, compounds23))$compound
```

<!-- CLR -->

```{r gh CLR, echo = F}
gh.joined.data <-
clean.gh.22 %>%
  dplyr::select(plant.id, compound, ngghr, swa, year, pair) %>%
  rbind(clean.gh.23 %>%
          dplyr::select(plant.id, compound, ngghr, swa, year, pair)) %>%
  distinct() %>%
  filter(compound %in% gh.compounds) %>%
  na.omit()

@@

### Box Plot

```{r gh boxplot comparison, echo = F, fig.align = 'center', fig.width = 10}
gh.joined.data.clr %>%
  # pivot to make all compounds in one column
  pivot_longer(-c(plant.id, year, swa, pair),
               names_to = "compound",
               values_to = "ngghr") %>%
  # ggplot code
  ggplot(aes(compound, ngghr, fill = swa)) +
  geom_boxplot(position = position_dodge(width = 0.9),
                 outlier.color = NA,
                 alpha = 0.5) +
  geom_jitter(aes(compound, ngghr),
              alpha = 0.3,
              position = position_jitterdodge(jitter.width = 0.2,
                                              dodge.width = .8),
              show.legend = FALSE,
              # pch = 21, 
              stroke = 0, 
              size = 1.5) +
  theme(axis.text.x = element_text(angle = 45, 
                                    hjust = 1),
        legend.background = element_rect(color = "black"),
        # legend.position = c(0.8,0.9),
        panel.background = element_rect(fill = "white"),
        axis.line = element_line(color = "black"),
        panel.grid.major.x = element_line(color = "#F2F2F2")) +
  labs(x = element_blank(),
       y = "VOC quantity (CLR(ng/g/hr))",
       title = "2022/23 Greenhouse Locoweed Volatile Profile Comparison") +
  scale_fill_discrete(name = "Endophyte",
                      labels = c("Not Present", "Present"))
```

::: callout-note
I chose to visualize the data using a box plot, not a bar plot, because our data has been CLR transformed. As we can see, there isn't a tremendous amount of obvious variance between E+ and E- samples. There is, perhaps, more of a difference at first glance than what we saw in our [field volatile analysis](field-locoweed-report.qmd#visual-comparisons) bar plots.
:::

### PCA

Principal Component Analysis (PCA) is a dimensionality reduction technique used to transform high-dimensional data into a lower-dimensional representation while preserving most of the original variability. It identifies the directions (principal components) that capture the maximum variance in the data and projects the data onto these components.

@@

## Check Assumptions

The statistical analyses we can perform are greatly impacted by our data's ability to meet statistical assumptions. One of the major assumptions here is normality. We are going to test this assumption using the `MVN` package.

@@

As we can see, our data violates the normality assumption. This is the case even though we ran it though a CLR transformation.
:::

::: callout-note
**A quick note here.**\
Many papers analyzing ecological or compositional data and VOC datasets often employ a statistical test known as [MANOVA](https://en.wikipedia.org/wiki/Multivariate_analysis_of_variance) (Multivariate Analysis of Variance). Despite the similarity of the data to ours in many instances, MANOVA is used even though the data being analyzed violates the normality assumption. Since our data clearly violates this assumption, we have decided against using this method.
:::

## Kruskal-Wallis Test

The [Kruskal-Wallis](https://en.wikipedia.org/wiki/Kruskal–Wallis_test) test is a non-parametric statistical test used to determine if there are differences between two or more independent groups. It assesses whether the distributions of scores across groups are the same or different, without assuming the data follow a normal distribution, making it suitable for ordinal or non-normally distributed data.

@@

### PERMANOVA

As described above, a [PERMANOVA](https://en.wikipedia.org/wiki/Permutational_analysis_of_variance) (Permutational Multivariate Analysis of Variance) uses permutation testing to generate many different rearrangements of the data and assess if the observed differences are statistically significant [@anderson2017]. This method is important as it considers all of the volatile profiles overall when generating a p-value.

@@

```{r PERMANOVA}
set.seed(1)
gh.permanova.output <-
adonis2(gh.permanova.data ~ gh.data.per.meta$swa, 
        strata = interaction(gh.data.per.meta$pair,
                             gh.data.per.meta$year),
        # method = "canberra",
        permutations = 9999,
        by = "terms")
```

```{r PERMANOVA table, echo = F}
kable(as.data.frame(gh.permanova.output),
      format = "html",
      caption = "Table 1: Greenhouse Locoweed VOC collection PERMANOVA results.   gh.data.per.meta$swa is our E+/E- variable.",
      align = "lccccc") %>%
  kable_styling(full_width = FALSE,
                bootstrap_options = c("striped", "bordered", "hover")) %>%
  row_spec(0, background = "#f2f2f9") 
```

@@

```{r random forest}
ranger_model <- ranger(swa ~ ., 
                       data = gh.joined.data.clr %>% 
                         dplyr::select(-plant.id),
                       importance = "impurity")
```

```{r random forest formatting and plot, echo = F}
var_importance <- importance(ranger_model)
